---
title: "Statistical Methods for Discrete Response, Time Series, and Panel Data (W271): Lab 2"
author: "Heather Feinstein, David Harding, Charlotte Swavola"
output: 
  pdf_document:
  toc: true
  number_sections: true
fontsize: 11pt
geometry: margin=1in
---
# Strategic Placement of Products in Grocery Stores

Answer **Question 12 of chapter 3 (on page 189 and 190)** of Bilder and Loughin's *"Analysis of Categorical Data with R"*.  Here is the background of this analysis, taken as an excerpt from this question:

In order to maximize sales, items within grocery stores are strategically placed to draw customer attention. This exercise examines one type of item:breakfast cereal. Typically, in large grocery stores, boxes of cereal are placed on sets of shelves located on one side of the aisle. By placing particular boxes of cereals on specific shelves, grocery stores may better attract customers to them. To investigate this further, a random sample of size 10 was taken from each of four shelves at a Dillons grocery store in Manhattan, KS. These data are given in the **cereal_dillons.csv** file. The response variable is the shelf number, which is numbered from bottom (1) to top (4), and the explanatory variables are the sugar, fat, and sodium content of the cereals.

```{r,message=FALSE,warning=FALSE}
# Load libraries
library(Hmisc)
library(MASS)
library(nnet)
library(stargazer)

# Load dataset
cereal <- read.csv("cereal_dillons.csv")
head(cereal)
# describe(cereal)
```

a. The explanatory variables need to be reformatted before proceeding further. 
    - First, divide each explanatory variable by its serving size to account for the different serving sizes among the cereals. 
  
    - Second, rescale each variable to be within 0 and 1.
```{r}
stand01 <- function(x) {(x - min(x)) /( max(x) - min(x))} 
cereal2 <- data.frame(Shelf = cereal$Shelf, sugar = 
                      stand01(x = cereal$sugar_g/cereal$size_g), 
                      fat = stand01(x = cereal$fat_g/cereal$size_g), 
                      sodium = stand01 (x = cereal$sodium_mg/cereal$size_g))

summary(cereal2)
```

   
  
b. Construct side-by-side box plots with dot plots overlaid for each of the explanatory variables. 
```{r}
par(mfrow=c(1,3))
boxplot(formula = sugar ~ Shelf, data = cereal2, ylab = "Sugar", xlab = "Shelf", 
        pars = list(outpch = NA),main="Weighted Sugar Content \nper Serving, by Shelf")
stripchart(x = cereal2$sugar ~ cereal2$Shelf, lwd = 2, col = "red", method = "jitter", 
           vertical = TRUE, pch = 1, add = TRUE,
           panel.first = grid(col = "gray", lty = "dotted"))

boxplot(formula = fat ~ Shelf, data = cereal2, ylab = "Fat", xlab = "Shelf", 
        pars = list(outpch = NA),main="Weighted Fat Content \nper Serving, by Shelf")
stripchart(x = cereal2$fat ~ cereal2$Shelf, lwd = 2, col = "blue", method = "jitter", 
           vertical = TRUE, pch = 1, add = TRUE,
           panel.first = grid(col = "gray", lty = "dotted"))

boxplot(formula = sodium ~ Shelf, data = cereal2, ylab = "Sodium", xlab = "Shelf", 
        pars = list(outpch = NA),main="Weighted Sodium Content \nper Serving, by Shelf")
stripchart(x = cereal2$sodium ~ cereal2$Shelf, lwd = 2, col = "green", method = "jitter", 
           vertical = TRUE, pch = 1, add = TRUE,
           panel.first = grid(col = "gray", lty = "dotted"))
```

```{r}
par(mar=c(2, 1, 2, 5),xpd = TRUE)
shelf.colors<-ifelse(test = cereal2$Shelf=="1", yes = "black", 
                    no = ifelse(test = cereal2$Shelf=="2", yes = "blue", no = ifelse(test = cereal2$Shelf=="3", yes = "green", no = "red")))
parcoord(x = cereal2[, c(2,3,4)], col = shelf.colors, var.label = TRUE)
title(main="Weighted Component Content by Shelf")
legend(3,0.5, legend = c("Shelf 4", "Shelf 3", "Shelf 2", "Shelf 1"), lty = "solid",
      col=c("red", "green", "blue", "black"), bty = 'n')                   
```

b. (cont'd) Discuss if possible content differences exist among the shelves.    
- Shelf 1 has a higher, narrow sodium distribution relative to the wide distributions of other components and the sodium distribution - for other shelves.     
- Shelf 2 has a higher, narrow distribution for sugar content, and hits the maximum fat content (though not for all samples).     
- Shelf 3 maintains a wide distribution across all three components. It has the samples with the lowest sugar, fat, and sodium content. 
- Shelf 4 has wide distributions for all components that are generally lower than the component distributions of other shelves. 
\newpage
c. The response has values of $1, 2, 3,$ and $4$. Under what setting would it be desirable to take into account ordinality. Do you think that this setting occurs here?

Ordinality could be considered for visibiility by a specific audience. For example, if you were examining kids' cereal specifically, the shelves would likely have the levels 2,1,3,4, for ordinality, as they are the most likely to be seen in that order by that specific audience. For fancy organic, perhaps 3,4,2,1- which corresponds to the eye levels of (most) adults. In this case, we haven't specified a cereal type or a particular audience to target, so an ordinality setting is not relevant.

d. Estimate a **multinomial regression model with linear forms of the sugar, fat, and sodium variables**. Perform **LRTs** to examine the importance of each explanatory variable.

```{r}
#Estimate model
mod.fit<-multinom(formula = Shelf ~ sugar + fat + sodium, data=cereal2)
summary(mod.fit)
```

```{r}
# LRT for sugar_g:
mod.fit.Ho_sugar<-multinom(formula = Shelf ~ fat + sodium, data=cereal2, trace=FALSE)
anova(mod.fit.Ho_sugar, mod.fit)  
```
The p-value for the likelihood ratio test is less than 0.05, therefore we can reject the null hypothesis for sugar, concluding that sugar is significant to the model.


```{r}
# LRT for fat_g:
mod.fit.Ho_fat<-multinom(formula = Shelf ~ sugar + sodium, data=cereal2, trace=FALSE)
anova(mod.fit.Ho_fat, mod.fit)   
```
We cannot reject the null hypothesis that fat is not significant to the model.


```{r}
# LRT for sodium_mg:
mod.fit.Ho_sodium<-multinom(formula = Shelf ~ sugar + fat, data=cereal2, trace=FALSE)
anova(mod.fit.Ho_sodium, mod.fit)   
```
For sodium, the p-value is less than 0.05. We can reject the null hypothesis and conclude that sodium is significant to the model

\newpage
e. Show that there are no significant interactions among the explanatory variables (including an interaction among all three variables).

```{r}
# LRT for sugar*sodium interaction term:
mod.fit.Ho_sugar_sodium <- multinom(formula = Shelf ~ sugar + fat + sodium + sugar:sodium, data=cereal2, maxit=200, trace=FALSE)
anova(mod.fit.Ho_sugar_sodium, mod.fit)   

# LRT for sugar*fat interaction term:
mod.fit.Ho_sugar_fat <- multinom(formula = Shelf ~ sugar + fat + sodium + sugar:fat, data=cereal2, maxit=200, trace=FALSE)
anova(mod.fit.Ho_sugar_fat, mod.fit)  

# LRT for fat*sodium interaction term:
mod.fit.Ho_fat_sodium <- multinom(formula = Shelf ~ sugar + fat + sodium + sugar:sodium, data=cereal2, maxit=200, trace=FALSE)
anova(mod.fit.Ho_fat_sodium, mod.fit)  

# LRT for sugar*fat*sodium interaction term:
mod.fit.Ho_sugar_fat_sodium <- multinom(formula = Shelf ~ sugar + fat + sodium + sugar:fat:sodium, data=cereal2, maxit=200, trace=FALSE)
anova(mod.fit.Ho_sugar_fat_sodium, mod.fit)  
```
We cannot reject any of the null hypotheses for the possible interaction coefficients. These interaction coefficients are not significant to the model.

f. Kellogg's Apple Jacks (http://www.applejacks.com) is a cereal marketed toward children. For a serving size of $28$ grams, its sugar content is $12$ grams, fat content is $0.5$ grams, and sodium content is $130$ milligrams. Estimate the shelf probabilities for Apple Jacks.

```{r}
# Data for Apple Jacks standardized
stand01.spec <- function(w,x) {(w - min(x)) /( max(x) - min(x))} 
newdata <- data.frame(sugar = stand01.spec(w = 12/28, x = cereal$sugar_g/cereal$size_g), 
                      fat = stand01.spec(w = 0.5/28, x = cereal$fat_g/cereal$size_g), 
                      sodium = stand01.spec(w = 130/28, x = cereal$sodium_mg/cereal$size_g))
newdata

# pi^
pi.hat<-predict(object = mod.fit, newdata = newdata, type = "probs")
round(pi.hat, 2)
```

g. Construct a plot similar to **Figure 3.3** where the estimated probability for a shelf is on the *y-axis* and the sugar content is on the *x-axis*. Use the mean overall fat and sodium content as the corresponding variable values in the model. Interpret the plot with respect to sugar content.
```{r}
curve(expr = predict(object = mod.fit, newdata = data.frame(sugar = x,
      fat = mean(cereal2$fat), sodium = mean(cereal2$sodium)), type = "probs")[,1], 
      main= expression(Shelf~hat(pi)~"vs Sugar content"),
      ylab = expression(Shelf~hat(pi)), xlab = "Sugar", xlim = c(min(cereal2$sugar), max(cereal2$sugar)), 
      ylim = c(0,1), col = "black", lty = "solid", lwd = 2, n = 1000,
      panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = predict(object = mod.fit, newdata = data.frame(sugar = x,
      fat = mean(cereal2$fat), sodium = mean(cereal2$sodium)), type = "probs")[,2], 
      col = "blue", lty = "solid", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = predict(object = mod.fit, newdata = data.frame(sugar = x,
      fat = mean(cereal2$fat), sodium = mean(cereal2$sodium)), type = "probs")[,3], 
      col = "green", lty = "solid", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = predict(object = mod.fit, newdata = data.frame(sugar = x,
      fat = mean(cereal2$fat), sodium = mean(cereal2$sodium)), type = "probs")[,4], 
      col = "red", lty = "solid", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

legend(x=0.2,y=1, legend=c("Shelf 4","Shelf 3","Shelf 2", "Shelf 1"), lty=c("solid"), col=c("red","green", "blue", "black"), bty="n", lwd = c(2,2,2))

```

```{r}
par(mfrow=c(1,2))
curve(expr = predict(object = mod.fit, newdata = data.frame(sugar = mean(cereal2$sugar),
      fat =x, sodium = mean(cereal2$sodium)), type = "probs")[,1],
      main= expression(Shelf~hat(pi)~"vs Fat"),
      ylab = expression(Shelf~hat(pi)), xlab = "Fat", xlim = c(min(cereal2$fat), max(cereal2$fat)), 
      ylim = c(0,1), col = "black", lty = "solid", lwd = 2, n = 1000,
      panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = predict(object = mod.fit, newdata = data.frame(sugar = mean(cereal2$sugar),
      fat =x, sodium = mean(cereal2$sodium)), type = "probs")[,2], 
      col = "blue", lty = "solid", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = predict(object = mod.fit, newdata = data.frame(sugar = mean(cereal2$sugar),
      fat =x, sodium = mean(cereal2$sodium)), type = "probs")[,3], 
      col = "green", lty = "solid", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = predict(object = mod.fit, newdata = data.frame(sugar = mean(cereal2$sugar),
      fat =x, sodium = mean(cereal2$sodium)), type = "probs")[,4], 
      col = "red", lty = "solid", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

legend(x=0,y=1, legend=c("Shelf 4","Shelf 3","Shelf 2", "Shelf 1"), lty=c("solid"), col=c("red","green", "blue", "black"), bty="n", lwd = c(2,2,2))

curve(expr = predict(object = mod.fit, newdata = data.frame(sugar = mean(cereal2$sugar),
      fat = mean(cereal2$fat), sodium = x), type = "probs")[,1],
      main= expression(Shelf~hat(pi)~"vs Sodium"),
      ylab = expression(Shelf~hat(pi)), xlab = "Sodium", xlim = c(min(cereal2$sodium), max(cereal2$sodium)), 
      ylim = c(0,1), col = "black", lty = "solid", lwd = 2, n = 1000,
      panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = predict(object = mod.fit, newdata = data.frame(sugar = mean(cereal2$sugar),
      fat =x, sodium = mean(cereal2$sodium)), type = "probs")[,2], 
      col = "blue", lty = "solid", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = predict(object = mod.fit, newdata = data.frame(sugar = mean(cereal2$sugar),
      fat =x, sodium = mean(cereal2$sodium)), type = "probs")[,3], 
      col = "green", lty = "solid", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = predict(object = mod.fit, newdata = data.frame(sugar = mean(cereal2$sugar),
      fat =x, sodium = mean(cereal2$sodium)), type = "probs")[,4], 
      col = "red", lty = "solid", lwd = 2, n = 1000,
      add = TRUE, panel.first = grid(col = "gray", lty = "dotted"))

legend(x=0,y=1, legend=c("Shelf 4","Shelf 3","Shelf 2", "Shelf 1"), lty=c("solid"), col=c("red","green", "blue", "black"), bty="n", lwd = c(2,2,2))

```

\newpage
h. Estimate odds ratios and calculate corresponding confidence intervals for each explanatory variable. Relate your interpretations back to the plots constructed for this exercise.
```{r,message=FALSE,warning=FALSE}
# Information about each variable to help with choosing c. Leave out Shelf column
sd.cereal2<-apply(X = cereal2[,-c(1)], MARGIN = 2, FUN = sd)
# sd.cereal2

#convert sd into (g/serving) units for interpretation. 0-1 are percentages of the overall range for g/serving
sd_convert<-function(sd,df_column,df_serving=cereal$size_g){
  var_range<-max(df_column/df_serving)-min(df_column/df_serving)
  return(sd*var_range)
}


c.value<-c(1, sd.cereal2)  # class = 1 is first value
c.value<-c.value[2:4] # drop intercept from c.value
round(c.value,2)

units<-c(g_serving=sd_convert(c.value[1],cereal$sugar_g),g_serving=sd_convert(c.value[2],cereal$fat_g),mg_serving=sd_convert(c.value[3],cereal$sodium_mg))
round(units,2)

# beta.hat_jr for r = 1, 2, 3  and j = 2, 3, 4
beta.hat2<-coefficients(mod.fit)[1,2:4]
beta.hat3<-coefficients(mod.fit)[2,2:4]
beta.hat4<-coefficients(mod.fit)[3,2:4]
```


```{r}
# Odds ratios for j = 2 vs. j = 1 
OR2_1<-exp(c.value*beta.hat2)
OR1_2<-1/exp(c.value*beta.hat2)

# Odds ratios for j = 3 vs. j = 2 
OR3_2<-exp(c.value*beta.hat3)
OR2_3<-1/exp(c.value*beta.hat3)

# for j = 3 vs j = 1
OR3_1<-OR3_2*OR2_1
OR1_3<-1/OR3_1

# Odds ratios for j = 4 vs. j = 3 
OR4_3<-exp(c.value*beta.hat4)
OR3_4<-1/exp(c.value*beta.hat4)

# for j = 4 vs j = 1
OR4_1<-OR4_3*OR3_1
OR1_4<-1/OR4_1 

# for j = 4 vs j = 2
OR4_2<-OR4_3*OR3_2
OR2_4<-1/OR4_2
```

```{r}
#build dataframes
OR_base=data.frame(OR2_1=round(OR2_1,2),
                    OR3_1=round(OR3_1,2),
                    OR4_1=round(OR4_1,2),
                    "-"=c("-","-","-"),
                    OR1_2=round(OR1_2,2),
                    OR3_2=round(OR3_2,2),
                    OR4_2=round(OR4_2,2),
                    "-"=c("-","-","-"),
                    OR1_3=round(OR1_3,2),
                    OR2_3=round(OR2_3,2),
                    OR4_3=round(OR4_3,2),
                    "-"=c("-","-","-"),
                    OR1_4=round(OR1_4,2),
                    OR2_4=round(OR2_4,2),
                    OR3_4=round(OR3_4,2))
```

```{r}
OR_base
```


```{r}                       
# Wald CIs
conf.beta<-confint(object = mod.fit, level = 0.95)
# round(conf.beta,2)  # Results are stored in a 3D array
# conf.beta[2:4,1:2,1]  # C.I.s for beta_2r
# conf.beta[2:4,1:2,2]  # C.I.s for beta_3r
# conf.beta[2:4,1:2,3]  # C.I.s for beta_4r
#CI for probability based on variable entry

# CIs for OR
ci.OR2<-exp(c.value*conf.beta[2:4,1:2,1])
ci.OR3<-exp(c.value*conf.beta[2:4,1:2,2])  
ci.OR4<-exp(c.value*conf.beta[2:4,1:2,3]) 

"Shelf 2,3,4 vs Shelf 1"
round(data.frame(low = ci.OR2[,1], up = ci.OR2[,2]), 2) #RELATIVE TO SHELF 1
round(data.frame(low = ci.OR3[,1], up = ci.OR3[,2]), 2) 
round(data.frame(low = ci.OR4[,1], up = ci.OR4[,2]), 2) 

"Shelf 3 vs Shelf 2"
round(data.frame(low = ci.OR3[,1]/ci.OR2[,1], up = ci.OR3[,2]/ci.OR2[,2]), 2) #shelf 3 relative to 2

"Shelf 4 vs Shelf 3"
round(data.frame(low = ci.OR4[,1]/ci.OR3[,1], up = ci.OR4[,2]/ci.OR3[,2]), 2) #shelf 4 relative to 3

```

**Odds ratio interpretations:**
The $c$ values (in $e^{c\beta_{jr}}$) used to evaluate the odds ratios for each component were calculated as 1 standard deviation unit. In grams per serving, these $c$ values are 0.15 g/serving of sugar, 0.03 g/serv fat, and 2.46 mg/serv sodium. 

Evaluating sugar first, for each c-value increase in sugar, the odds that the cereal would be on shelf 3 are 21.48 times as large as the odds it would be on shelf 4, 12.98 times as large that it'd be on shelf 1 as shelf 3, and 2.06 times as large that it would be on shelf 2 as shelf 1. This matches the $\hat{\pi}$ graph for sugar as shelves 1 and 2 have the highest probabilities for increased sugar content, however doesn't match the relatioship between 3 and 4. This discrepancy may be a result of a smoothing error, as shelf 3 has an overall wider range of sugar values, including some at higher sugar levels as the odds ratios suggest, but also has more samples at lower sugar values than shelf 4. Incorporating the confidence intervals, the CI of the odds ratio that the cereal will appear on shelf 2 vs shelf 1 includes 1, so we cannot definitively say the odds the cereal will be on shelf 2 are larger than shelf 1. We can, however, say that the odds the cereal will be on shelves 3 or 4 relative to shelf 1 are less than 1 for each unit increase in sugar.     

Looking at fat, again the odds that a cereal will be on shelf 2 are at least twice as large as the odds of the other shelves. The $\hat{\pi}$ graphs echo this, for as the fat content increases, the likelihood the cereal will appear 2 increases while all others decrease. Again, The odds ratio confidence intervals span 1 for shelf 2 vs shelf 1,  however are entirely below 1 for shelf 3 vs shelf 2 and shelf 4 vs shelf 3, indicating that the odds of the cereal to be on shelf 2 are larger than shelf 3 or 4 for an increase in fat. 

Now, the odds ratios for sodium reach extreme numbers, especially relative to shelf 1. The behavior of the $/hat{/pi}$ vs sodium content for shelf 1 increases from 0 to 1 in about 0.3 units, almost the same as our c-value, potentially because only high-sodium samples appear on shelf 1. This is reiterated by the confidence intervals, as all the other shelf intervals are below 0.5 relative to shelf 1. This means that for each unit increase in sodium, the odds that a cereal appears on any other shelf are at most 1/2 as large as the odds that the cereal appears on shelf 1. It's important to note that the $\hat{\pi}$ graph for sodium actually exceeds the total probability boundary of 1. This indicates an instability in the model. 








